Rust for Scientific Computing

### 2025-08-22 main.tex
include figures in text
add first list of crates
- still incomplete
- probably redundant
- contains some wrongly formatted results

@@

\subsection{High-Performance \& Scalable Computing}
\subsubsection{Parallel Computing}
(rayon)

\subsubsection{GPU \& Accelerators}
(rust-cuda, wgpu~\cite{Fitzgerald2025})

@@

\subsubsection{Debugging}
(Rust Errors as values, tracing, logging) ecosystem

@@

@@

>>>>

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{figures/stackoverflow-popular-languages.pdf}%
    \includegraphics[width=0.5\textwidth]{figures/pypi-org-used-languages.pdf}
    \caption{TODO}
\end{figure}

@@

\subsection{High-Performance \& Scalable Computing}
\subsubsection{Parallel Computing}
(rayon, compare to openmp, MPI bindings)

\subsubsection{GPU \& Accelerators}
(rust-cuda, cudarc, wgpu~\cite{Fitzgerald2025}, opencl, mention SPIRV)

@@

\subsection{Creature Comforts}
\subsubsection{Debugging}
(Rust Errors as values, tracing, logging) ecosystem

@@

\subsubsection{Bindings}
(pyo3, maturin, C, cxxbridge, cmake)

@@

%###################################################################################################
\pagebreak
\newcounter{supplementSection}
\renewcommand{\thesubsection}{S\arabic{supplementSection}}
\newcommand{\supplement}[1]{%
    \stepcounter{supplementSection}%
    \subsection{#1}
}
\renewcommand{\thesection}{}

\section{Supplementary Material}
\supplement{List of all crates}

\paragraph{Scientific Computing in Rust monthly Newsletter}

approx-derive\\
A procedural macro that automatically implements approximate equality traits for
floating-point types and custom structs. It builds on the approx crate to make testing numerical
code easier. This helps ensure tolerance-based equality checks are consistent and less error-prone.

approxim\\
A library for approximation algorithms in numerical computing. It provides tools for
approximating functions, datasets, or computations efficiently. The focus is on balancing precision
with performance in scientific workflows.

bacon-sci\\
A scientific computing library built on Rust that provides a foundation for numerical
and algebraic operations. It aims to be a lightweight alternative to larger ecosystems. Its design
emphasizes safety, performance, and modularity.

cargo-upgrades\\
A Cargo subcommand for upgrading dependencies in a Rust project. It automates
version checking and updates Cargo.toml accordingly. This reduces the manual overhead of keeping
scientific projects up to date.

EGObox\\
A library for evolutionary game theory and agent-based modeling. It allows researchers to
explore strategies, equilibria, and dynamics in multi-agent systems. Designed with flexibility to
model diverse game-theoretic scenarios.

faer\\
A linear algebra library for Rust with focus on high-performance matrix operations. It
provides dense and sparse routines with attention to numerical stability. The crate aims to become
a foundation for scientific and machine-learning applications in Rust.

fact.rs\\
A crate for fast matrix factorizations and related linear algebra decompositions. It
provides algorithms like LU, QR, and eigenvalue methods. Its goal is to complement crates like faer
with specialized routines.

globalsearch\\
A Rust crate for global optimization and search algorithms. It supports solving
complex optimization problems where local minima are common. Useful in scientific simulations and
computational experiments.

indexing\_fmt\\
A utility crate for formatting indices and ranges in scientific computing contexts.
It makes working with slices and multi-dimensional arrays more ergonomic. This helps in debugging
and presenting results.

integraal\\
A symbolic and numerical integration library for Rust. It supports computing definite
and indefinite integrals for a variety of functions. Its goal is to provide a reliable integration
toolkit for researchers.

kifmm-rs\\
A Rust implementation of the kernel-independent fast multipole method (FMM). It
accelerates particle simulations and large-scale interactions in physics. The focus is on
high-performance computing and scalability.

ndgrid\\
A crate for generating multi-dimensional grids, similar to MATLAB’s ndgrid. It is
especially useful in scientific computing for parameter sweeps and simulations. Provides ergonomic
grid construction for Rust’s array ecosystem.

ndelement\\
A crate for manipulating individual elements in multi-dimensional arrays. It offers
convenient indexing and iteration across dimensions. Designed to complement libraries like ndarray.

ndarray\\
A core Rust crate for n-dimensional arrays, inspired by NumPy. It provides efficient
array structures, broadcasting, slicing, and linear algebra support. Widely used as a foundation
for many Rust scientific computing libraries.

ploc\\
A parallelized algorithm library for localization and clustering tasks. Focuses on scalable
implementations of clustering approaches. Intended for large datasets in scientific applications.

Polars\\
A fast DataFrame library for Rust, with APIs also available in Python. It provides
powerful data manipulation, joins, and group-by operations with performance close to Apache Arrow.
Designed for data science, analytics, and scientific workflows.

Quant-Iron\\
A crate for quantitative finance and scientific modeling. It implements financial
instruments, stochastic processes, and numerical solvers. Designed to bridge scientific computing
with financial applications.

paradis\\
A discrete dislocation dynamics simulator written in Rust. It is used in materials
science to model dislocation behavior under stress. The library emphasizes performance and
scalability for large simulations.

repgenerate\\
A tool for generating reproducible reports from computational experiments. It
integrates with Rust code to ensure scientific workflows can be repeated. Useful for open and
transparent research.

rsmpi\\
Rust bindings to the Message Passing Interface (MPI). It enables parallel computing across
distributed systems, a key component in HPC. Provides safe abstractions while maintaining low-level
control.

serde\\
A widely used Rust serialization and deserialization framework. It supports many formats
like JSON, YAML, and custom binary encodings. In scientific computing, it helps with data exchange
and configuration management.

sequenceprofiler\\
A bioinformatics library for profiling biological sequences. Provides tools for
analyzing DNA, RNA, or protein sequences. Aimed at high-performance sequence analysis in Rust.

zarrs\\
A Rust implementation of the Zarr data format for chunked, compressed N-dimensional arrays.
It allows efficient storage and manipulation of large scientific datasets. Especially useful in
machine learning and climate science.

\paragraph{JOSS}

fastatomstruct\\
A Rust library for high-performance analysis of atomic systems. It provides tools
for structural and dynamical computations, enabling simulations and data analysis in computational
chemistry and materials science. Optimized for speed, it can handle large datasets efficiently.

grepq\\
A Rust tool designed to process and filter FASTQ files efficiently. It matches sequences
against sets of regular expressions, making it useful for high-throughput sequencing data analysis.
Its focus is on speed and flexibility for genomic workflows.

pengWann\\
A library that computes chemical bonding descriptors using Wannier functions. It aids in
understanding electronic structure and bonding characteristics in materials. Suitable for quantum
chemistry and materials science applications.

wrenfold\\
A Rust library for symbolic computation and code generation tailored to robotics
applications. It allows automatic generation of control and kinematics code from symbolic
expressions. Supports efficient development of robotic algorithms.

kifmm-rs\\
A Rust framework implementing the kernel-independent fast multipole method (FMM). Used
to accelerate N-body computations in physics and engineering. Optimized for performance and
scalable to large simulations.

startinpy\\
A Python package for modeling and processing 2.5D terrains. Supports triangulated
meshes and geospatial analysis. Useful in geoscience, terrain analysis, and visualization.

Back to sequences\\
A tool for tracing the origins of k-mers in sequencing data. Helps analyze and
understand sequence composition and frequency. Useful for genomics and bioinformatics research.

sourmash v4\\
A bioinformatics toolkit for genome and metagenome analysis. Provides fast
comparison, search, and analysis of sequencing data. Supports scalable workflows with large
datasets.

extendr\\
A Rust library that facilitates seamless integration with the R programming language.
Allows calling Rust functions from R and vice versa. Focuses on performance and ease of use for
scientific computing.

Pywaterflood\\
A Python library for analyzing well connectivity in hydrogeology or petroleum
engineering. Uses capacitance-resistance modeling to simulate flow and connectivity. Supports
data-driven analysis for reservoir studies.

Gibbs Sea Water Oceanographic Toolbox\\
A Rust implementation of the TEOS-10 toolbox for
oceanographic computations. Provides tools for thermodynamic and salinity calculations in seawater.
Useful for oceanographers and climate researchers.

Raphtory\\
A toolkit for building and analyzing temporal graphs. Supports both Rust and Python
interfaces. Enables high-performance analytics on time-evolving network data.

HDT\\
A Rust library for working with HDT, a binary compression format for RDF data. Optimizes
storage and query of large knowledge graphs. Useful for semantic web and linked data applications.

egobox\\
A Rust toolbox for performing global optimization. Implements algorithms for efficiently
exploring high-dimensional spaces. Useful in engineering, machine learning, and simulation-based
optimization.

Fast k-medoids Clustering\\
A high-performance implementation of k-medoids clustering in Rust, with
Python bindings. Supports scalable clustering of large datasets. Ideal for machine learning and
data analysis tasks.

FEM\_2D\\
A Rust library for 2D finite element simulations. Supports hp-refinement for adaptive
mesh accuracy. Designed for engineering, physics, and numerical analysis applications.

retworkx\\
A Python graph library with Rust backend for performance. Supports graph creation,
traversal, and analysis. Useful in network science, bioinformatics, and algorithm development.

Rasusa\\
A bioinformatics tool for subsampling sequencing reads. Allows control of coverage in
genomic datasets. Helps optimize computational pipelines and reduce data size for analysis.

Sepia\\
A Rust tool for classifying sequencing reads by taxonomy. Provides high-performance
analysis for metagenomics studies. Helps researchers quickly categorize large datasets.

RustBCA\\
A Rust library implementing the binary collision approximation (BCA) for ion-material
simulations. Enables efficient modeling of particle interactions in materials. Used in physics and
materials research.

City2BA\\
A Rust package for generating synthetic bundle adjustment datasets. Useful in computer
vision and photogrammetry. Helps test and benchmark 3D reconstruction algorithms.

\paragraph{Scientific Computing in Rust 2025 talks}

Diffsol\\
A crate for solving differential equations.

Pixi\\
The missing companion to Cargo.

ploc\\
Efficient point location with ploc.

NPB-Rust\\
NAS parallel benchmarks in Rust.

cellular\_raza\\
Cellular agent-based modeling from a clean slate.

ninterp\\
Numerical interpolation in N-dimensions.

Arrow\\
Arrow, Rust, and cross-language data science tooling.

Juice\\
Juice your simulations: what science can learn from game development.

eVaiutilities\\
Research data management post eVai.

Honeycomb\\
Combinatorial maps implementation for meshing applications.

Stochastic\\
Stochastic calculus in Rust.

Exponential\\
Exponential time integration for stiff systems in Rust.

Rust\\
Rust for chip design algorithms and EDA software.

Using\\
Using Rust to mitigate breaking changes in scientific software.

The\\
The pleasure and pain of developing cross-platform CPU vector code in Rust.

Deimos\\
Open-source scientific data acquisition \& laboratory controls.

Rust\\
Rust is RAD and this is why.

Diffsol\\
A crate for solving differential equations.

Pixi\\
The missing companion to Cargo.

ploc\\
Efficient point location with ploc.

\paragraph{Scientific Computing in Rust 2024 talks}

AcoDyn\\
Fluid dynamics on the GPU powered by Rust.

RLST\\
Linear algebra frameworks in Rust for the solution of partial differential equations.

Raphtory\\
Temporal graphs from Scala to Rust.

Bless\\
Transparently logging program outputs.

sophus-rs\\
A fast electromagnetics library with Rust and Python.

sophus-rs\\
A fast electromagnetics library with Rust and Python.

Rootfinders\\
Rootfinders for Rust.

Rstats\\
Multidimensional data analysis in Rust.

extendr\\
Frictionless bindings for R and Rust.

Rusting\\
Rusting RSRS.

Physics\\
Physics simulations in Bevy.

An\\
An embedded memory cache for persistent memory access.

Polars\\
Polars in the factory: when databases trump collections for modeling.

Perpetaul\\
Perpetaul: a hyperparameter-free gradient boosting machine.

Writing\\
Writing maths software in Rust after 20 years of C++: graph isomorphism and groups.

Productive\\
Productive and reliable constitutive modeling in Rust.

Efficient\\
Efficient 1-bit matrix approximations.

Developing\\
Developing Rust-based R packages using the Roxido framework.

Rusph\\
Rusph: a SPH astrophysical simulation code in the Rust programming language.

Back\\
Back to basics: rebuilding computational biomechanics crate by crate.

Cargo\\
Cargo +GPU build: An early outlook.

Writing\\
Writing a grid library for finite and boundary element methods.

\paragraph{Scientific Computing in Rust 2023 talks}

jiro\_nn\\
Low-friction high-detail Deep Learning framework in Rust

\url{https://chrisproject.org/}\\
ChRIS is an open-source platform for computational research and medicine.

struqture\\
Represent quantum mechanical operators, Hamiltonians and open quantum systems.
\url{https://quantumsimulations.de/}

salso, caviarpd, \& fangs\\
R extensions written in Rust

faer-rs\\
A linear algebra foundation for the Rust programming language.

Lamellar\\
A Rust-based asynchronous tasking and PGAS runtime for high-performance computing.

Sage\\
Enables fast proteomics searching and quantification at scale.

Bernstein–Bézier\\
Finite elements for RMM in Rust. ResearchGate

Roqoqo\\
A quantum computing toolkit in Rust.

Lace\\
Bayesian tabular data analysis for Rust (and Python).

Rudolph\\
Improving productivity of medical research with Rust CLI tools.

Rstats\\
Multidimensional data analysis in Rust.

extendr\\
Frictionless bindings for R and Rust.

Rusting\\
Rusting RSRS.

Physics\\
Physics simulations in Bevy.

An\\
An embedded memory cache for persistent memory access.

Polars\\
Polars in the factory: when databases trump collections for modeling.

Perpetaul\\
Perpetaul: a hyperparameter-free gradient boosting machine.

Writing\\
Writing maths software in Rust after 20 years of C++: graph isomorphism and groups.

Productive\\
Productive and reliable constitutive modeling in Rust.

Efficient\\
Efficient 1-bit matrix approximations.

Developing\\
Developing Rust-based R packages using the Roxido framework.

Rusph\\
Rusph: a SPH astrophysical simulation code in the Rust programming language.

Back\\
Back to basics: rebuilding computational biomechanics crate by crate.

Cargo\ \ Cargo +GPU build: An early outlook.

Writing\\
Writing a grid library for finite and boundary element methods.

\paragraph{FAANG Companies}

Amazon (AWS)\\
AWS developed Firecracker, a microVM monitoring service powering AWS Lambda and AWS Fargate.
Firecracker is built primarily in Rust to offer fast cold boot times and secure workload isolation.

Rust-supported runtimes like AWS Lambda also allow developers to write serverless functions using
Rust. DEV Community Litslink

Google\\
Fuchsia, Google's modern operating system, uses Rust extensively for system components such as
drivers and services to enhance security and reliability.

Android integration: Rust has been adopted within the Android Open Source Project (AOSP) to improve
memory safety in critical system components. DEV Community Litslink

Meta (Facebook)\\
Mononoke, Facebook’s large-scale source control backend for Mercurial, was written in Rust to
improve performance and reliability in its monorepo.

Facebook employs Rust in developing CLI tools, distributed systems, and cron jobs. Engineers write
millions of lines of Rust internally. Engineering at Meta Facebook Developers

Apple\\
Apple’s Cloud Traffic team and backend infrastructure have been shifting from C to Rust, with new
functionality being primarily built in Rust. This change was driven by performance, security, and
developer efficiency goals. Reddit

Microsoft\\
Although not part of FAANG, it’s noteworthy—Microsoft is rewriting parts of the Windows kernel and
system components in Rust (e.g., DWriteCore, GDI Regions), prioritizing memory safety.

Azure services—like Azure IoT Edge—also incorporate Rust. Litslink Wikipedia
